Amplitude metrics for cellular \\ circadian bioluminescence reporters

### 2014-09-25 response_letter.tex

\begin{document}
\today\\

{\itshape Francis J. Doyle III}\\
Dept.\ of Chemical Engineering\\
Univ.\ of California, Santa Barbara\\
Santa Barbara, CA 93106-5080\\

{\itshape Leah Edelstein-Keshet}\\
Editorial Board Member\\
Biophysical Journal\\

Thank you for your email on September 24th, 2014, inviting us to submit a revised version of our manuscript, ``Amplitude metrics for cellular circadian bioluminescence reporters'' (MS: 2014BIOPHYSJ304278R) by Peter C. St. John, Stephanie R. Taylor, John H. Abel, and Francis J. Doyle~III.

We thank the reviewers for their second round of comments of the manuscript, in particular Reviewer \#2 for their continued attention to detail. We believe addressing these remaining concerns and suggestions have improved the polish of the manuscript.

@@

Please see our detailed responses to the reviewers below:

\section*{Reviewer \#1}
\begin{reviewer}
The authors have largely responded you my concerns.

Typos and suggestions

- p. 2: "fail to capture the collective dynamics of a population oscillators" 
\end{reviewer}

This typo has been addressed, the sentence now reads:

\begin{manuscript}{Page 2}
  Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population {\bfseries of} oscillators.
\end{manuscript}

\begin{reviewer}
- mu is defined a few lines after it is first used in Eq. (20). Perhaps define it first? 
\end{reviewer}

This change has also been made.

\begin{manuscript}{Page 7}
To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times.
Since $x(t)$ approaches the reference as $t \to \infty$, {\bfseries the means of both trajectories are equal and can be calculated by
\begin{equation}
  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta.
  \tag{20}
\end{equation}
The amplitude change metric is defined as}
\begin{equation}
  \begin{aligned}
    \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
    &= \int_0^\infty h(t) \; dt.
  \end{aligned}
  \tag{21}
\end{equation}
\end{manuscript}

### 2014-09-25 main_text.tex
starting second revision, adding changes requested by first reviewer

The choice of modeling framework dictates the type of amplitude response which can be predicted.
Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population oscillators.
Likewise, phase-only models correctly capture the change in synchrony of a population, but do not capture fluctuations in amplitude of individual oscillators.
Explicit stochastic simulations of populations of cells are capable of realistically capturing both single-cell and population-level effects, and have been successfully used to understand the response of coupled oscillators to external VIP perturbation \cite{An2013}.
However, these methods are computationally expensive, and cannot be used for an analytical understanding of amplitude response.

@@

Ordinary differential equation (ODE) models take the general form
\begin{equation}
  \frac{dx}{dt} = f(x(t), p)
  \label{eq:odefn}
\end{equation}
in which $x(t)$ represents the concentrations of the state variables, such as mRNA and protein concentrations, $f$ contains information on the production, degradation, and reactivity of the states, and $p$ are the kinetic parameters which govern reaction kinetics.
Limit cycle models are ODE models in which the solution approaches a steady state oscillatory trajectory, satisfying the equation: \begin{equation}
  \lim_{t \to \infty} \left[ x(t + T) - x(t) \right] = 0
  \label{eq:limit}
\end{equation}
The period is the smallest $T > 0$ for which Eq.~\ref{eq:limit} holds.
The points on the stable limit cycle are denoted by $x^\gamma(\theta)$ with each point assigned to a value of a phase variable $\theta \in [0, 2\pi)$.
For convenience, time in Eq.~\ref{eq:odefn} can be rescaled such that the period is $2\pi$:
\begin{equation}
  \tilde{t} = \frac{2\pi}{T}t; \quad \tilde{f} = \frac{T}{2\pi}\;f; \quad \frac{dx}{d\tilde{t}} = \tilde{f}(x(\tilde{t}), p)
  \label{eq:that}
\end{equation}
The phase variable $\theta$ is therefore defined on the limit cycle as $\theta = \tilde{t}\mod 2\pi$, with $\theta = 0$ assigned to a unique and identifiable point.

@@

In this study we restrict our analysis to temporary perturbations capable of entraining an oscillatory system, excluding permanent parameter changes arising, for instance, from genetic knockout experiments.
The most simple entraining perturbation involves adding or removing components to a limit cycle system, resulting in a perturbed trajectory $x(t)$.
Here the initial conditions are determined by the strength of the perturbation and the phase at which it is applied:
\begin{equation}
  x(0) \coloneqq x^\gamma(\theta_0) + \Delta x(0)
  \label{eq:stateperturbation}
\end{equation}
This trajectory evolves according to Eq.~\ref{eq:odefn}, eventually returning to $x^\gamma$.
It is useful to express this trajectory by the deviation from the limit cycle:
\begin{equation}
  \Delta x(\tilde{t}) = x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0)
  \label{eq:delxt}
\end{equation}

In addition to perturbations directly to the state of the system, oscillators can also be perturbed by temporary changes to the parameters.
For a parameter pulse $\Delta p$ of duration $\tilde{d}$ (in radians) that ends at $\theta_0$, the oscillator deviates from the limit cycle trajectory according to:
\begin{align}
  x_{\tilde{d}}(-\tilde{d}) &= x^\gamma(\theta_0 - \tilde{d}) \\
  \frac{dx_{\tilde{d}}}{d\tilde{t}} &= \tilde{f}(x_{\tilde{d}}(\tilde{t}), p + \Delta p) \label{eq:ode_pert}
\end{align}
The pulse trajectory $x_{\tilde{d}}(\tilde{t})$ is then integrated from $\tilde{t} = -\tilde{d} \to 0$, at which point the pulse is removed.
A temporary parameter pulse is therefore equivalent to a state perturbation, but with the perturbation at $t = 0$ defined by:
\begin{equation}
  \Delta x(0) = x_{\tilde{d}}(0) - x^\gamma(\theta_0)
\end{equation}
A schematic depicting an example perturbed and reference trajectory is shown in Fig. 1A.

The response of the system to very small perturbations is more efficiently calculated using ODE sensitivity analysis \cite{Rabitz1983}.
The sensitivity matrix:
\begin{equation}
  S(t) = \frac{dx(t)}{dx(0)} = \lim_{\Delta x(0) \to 0}\frac{\Delta x(t)}{\Delta x(0)}
  \label{eq:senslimit}
\end{equation}
can be calculated directly by integrating:
\begin{equation}
  \frac{d}{dt} S(t)  = \frac{df(x(t),p)}{dx}\; S(t)
  \label{eq:odesens}
\end{equation}
with $S(0) = I$.

\subsection*{Phase response curves}
Quantifying phase changes following a perturbation has been well studied and is particularly relevant for models of circadian rhythms \cite{Kramer1984, Taylor2008a}.
A phase response curve (PRC) maps the change in phase resulting from the same perturbation applied at each initial phase.
Infinitesimal PRCs, the derivative of the phase change with respect to the perturbation, can be defined for state and parameter-impulse perturbations \cite{Taylor2008a}.
Methods for efficiently calculating these quantities using ODE sensitivity analysis have been developed \cite{Taylor2008a}, with the important result that the parameter- and state-impulse PRCs can be related by the Jacobian matrix:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp} 
  \label{eq:pPRCequiv}
\end{equation}
This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.

\subsection*{Phase-diffusion model}
Large populations of oscillators are typically described using phase-only models \cite{Rougemont2006}, in which the state of each oscillator is represented only by its phase, $\theta$.
The synchrony of the population can be modeled using a probability density function $p(\theta, \tilde{t})$ that describes the probability of finding an oscillator at each phase \cite{Kuramoto1984}.
The usefulness of probability density functions in describing the phase and amplitude responses of populations of circadian cells has been previously shown \cite{Ukai2007}.
As with all probability density functions:
\begin{equation}
  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1
\end{equation}
The shape of $p(\theta, \tilde{t})$ changes as the cells advance in time.
Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population \cite{Teramae2004}.
For a infinite population of oscillators, these effects are well-described by a Fokker-Plank equation \cite{Stein1965}:
\begin{equation}
  \frac{\partial p}{\partial \tilde{t}} = \frac{\partial p}{\partial \theta} + d\frac{\partial^2 p}{\partial \theta^2}\\
  \label{eq:pde}
\end{equation}
Due to the rescaling of $t$, the mean period of the population is $2\pi$.
Here, the $\nicefrac{\partial p}{\partial \theta}$ term, analogous to convection, describes the mean oscillatory period, while the $\nicefrac{\partial^2 p}{\partial \theta^2}$ term describes the diffusion of phases across $[0, 2\pi)$.
The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data \cite{Rougemont2007}.
Eq.~\ref{eq:pde} has periodic boundary conditions, with initial condition $\phi(\theta)$ as the phase population at $t=0$:
\begin{align}
  \text{BCs:}\quad p(0, \tilde{t}) &= p(2\pi, \tilde{t}) \\
  \frac{\partial p}{\partial \theta}(0, \tilde{t}) &= \frac{\partial p}{\partial \theta}(2\pi, \tilde{t}) \\
  \text{IC:}\quad p(\theta, 0) &= \phi(\theta)\label{eq:pde_ic}
\end{align}
The solution of Eqs.~\ref{eq:pde}-\ref{eq:pde_ic} is well-characterized, with $p(\theta, \tilde{t})$ evolving in time as the convolution of the initial conditions with a wrapped normal distribution with mean $\tilde{t}$ and standard deviation $\sqrt{2d\tilde{t}}$ \cite{Chirikjian2009}:
\begin{equation}
  p(\theta, \tilde{t}) = \phi(\theta) * \mathcal{WN}(\theta; \tilde{t},
  \sqrt{2\, d\, \tilde{t}})
\end{equation}
in which the wrapped normal distribution \cite{Mardia2009} is defined as:
\begin{equation}
  {\cal WN}(\theta; \mu, \sigma) =
  \frac{1}{\sigma\sqrt{2\pi}}\sum_{k=-\infty}^\infty \exp\left[\frac{-(\theta
  - \mu + 2\pi k)^2}{2\sigma^2}\right]
  \label{eq:wrapped_normal}
\end{equation}
Since the convolution of two normal distributions is also a normal distribution, it is efficient when possible to describe $\phi(\theta)$ as a normal distribution with mean $\mu_0$ and standard deviation $\sigma_0$, such that $p(\theta, \tilde{t})$ can be found analytically through:
\begin{equation}
  p(\theta, \tilde{t}) = \mathcal{WN}(\theta; \mu_0 + \tilde{t},
  \sqrt{\sigma_0^2 + 2d^2\tilde{t}^2})
\end{equation}

@@

Following a perturbation to a single cell, the path that the perturbed trajectory $x(t)$ takes in returning to the limit cycle will have a different amplitude than the unperturbed trajectory.
While such an amplitude change will only have a finite duration, it plays an important role when perturbations are repeatedly received by the clock, such as a peripheral oscillator entrained to daily metabolic stimuli.
To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times:
\begin{equation}
  \begin{aligned}
    \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
    &= \int_0^\infty h(t) \; dt
  \end{aligned}
  \label{eq:ampchangedefinition}
\end{equation}
This amplitude metric was chosen over alternatives, such as peak-trough distance, for its analytical tractability.
Since $x(t)$ approaches the reference as $t \to \infty$, the means of both trajectories are equal.
The mean of the limit cycle can therefore be calculated as:
\begin{equation}
  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta
  \label{eq:mu}
\end{equation}
The integrand in Eq.~\ref{eq:ampchangedefinition}, abbreviated by $h(t)$, compares the variances of the reference and perturbed trajectories.
When $h(t) > 0$, the trajectory is further from the mean than the reference, and similarly when $h(t) < 0$ the trajectory is closer to the mean than the reference.
Thus the overall amplitude change can be calculated by integrating $h(t)$ until the two trajectories converge, returning an amplitude value for each state variable.
Amplitude change for a two-dimensional oscillator is easy to visualize graphically.
In Fig.~1 B-D, the same state perturbation $\Delta x(0)$ applied at two different phases results in opposite amplitude changes, depending on whether the perturbation shifts the trajectory to the interior of the limit cycle (reduced amplitudes) or to the outside the limit cycle (increased amplitudes).
Trajectories for the second state variable, $Y(\tilde{t})$, and the corresponding integrand for the amplitude change equation, $h(\tilde{t})$, demonstrate how this transient change is quantified.

@@

We therefore derive an expression for the infinitesimal ARC, defined as:

@@

  \frac{dA}{dx} \coloneqq \lim_{\Delta x(0) \to 0} \frac{\Delta A\left(x(\tilde{t}),\ x^\gamma(\tilde{t} + \theta_0 + \Delta\theta)\right)}{\Delta x(0)} = \int_0^\infty \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} \; dt

@@

  O(\Delta\theta^2)

@@

Simplifying the integrand in Eq.~\ref{eq:sARC}:

@@

    2(x^\gamma(t_\theta) - \mu)\right) \right]

@@

  &= 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right)

@@

  \frac{dA}{dx} = \int_0^\infty 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right) \; dt

@@

Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:

@@

  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p}

@@

  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}

@@

We next briefly describe how to calculate the PRC and ARC for a phase-density model.
These approaches are commonly used in understanding phase models \cite{Kuramoto1984, Ukai2007}, and are presented here to match previous definitions for single cells.
A phase transition curve, $g(\theta) = \theta + \Delta\theta$, maps the phase of an oscillator before perturbation to its phase after the perturbation.
Since individual oscillators are neither created nor destroyed during the perturbation, it is possible - yet numerically difficult - to directly calculate the phase probability distribution following perturbation using the standard change of variables relation:
\begin{equation}
  \hat{p}(\theta, \tilde{t})\; dg(\theta) = p(\theta, \tilde{t})\; d\theta
  \label{eq:pdfinversion}
\end{equation}
However, it is easier to estimate the mean and standard deviation of the perturbed population $\hat{p}(\theta, \tilde{t})$ using directional statistics \cite{Mardia2009}.
A population defined on the unit circle can be described by a complex variable $z = \rho e^{i\bar{\theta}}$, where $\bar{\theta}$ is the mean phase and $\rho$, the synchronization index, is related to the standard deviation of the population.
For $\rho = 1$, the population is clustered about one mean phase, while for $\rho = 0$ the population is evenly balanced across the unit circle.
Complex variables for the population before and after perturbation can be calculated via:
\begin{align}
  z &\coloneqq \int_0^{2\pi} e^{i\theta} p(\theta, \tilde{t}) \; d\theta \label{eq:zbar}\\
  \hat{z} &\coloneqq  \int_0^{2\pi} e^{i\theta} \hat{p}(\theta, \tilde{t}) \; d\theta =
  \int_0^{2\pi} e^{i g(\theta)} p(\theta, \tilde{t}) \; d\theta
  \label{eq:zhat}
\end{align}
In Eq.~\ref{eq:zhat}, we avoid calculating the perturbed population explicitly by instead integrating over the new phases at the prior population density function.
Population-level amplitude and phase responses can therefore be calculated by:
\begin{align}
  \Delta \bar{\theta} &= \angle z - \angle \hat{z} \\
  \Delta \rho &= |z| - |\hat{z}|
  \label{eq:popampchange}
\end{align}
Population-level PRCs and ARCs can be tabulated by solving Eqs.~\ref{eq:zbar}-\ref{eq:popampchange} for populations $p(\theta, \tilde{t})$ with different mean phases.
It is important to note that the ARC at the population level strongly depends on the slope of the PRC, as has been shown previously \cite{Ukai2007}.

@@

Assuming each oscillator in the population follows the dynamics described by $x^\gamma(\theta)$, the unperturbed mean population-level expression, $\bar{x}(\tilde{t})$, can be found by taking the weighted average of the expression level over the current population:
\begin{equation}
  \bar{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) p(\theta, \tilde{t}) \;
  d\theta
  \label{eq:xbar}
\end{equation}

>>>>

The choice of modeling framework dictates the type of amplitude response which can be predicted.
Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population of oscillators.
Likewise, phase-only models correctly capture the change in synchrony of a population, but do not capture fluctuations in amplitude of individual oscillators.
Explicit stochastic simulations of populations of cells are capable of realistically capturing both single-cell and population-level effects, and have been successfully used to understand the response of coupled oscillators to external VIP perturbation \cite{An2013}.
However, these methods are computationally expensive, and cannot be used for an analytical understanding of amplitude response.

@@

Ordinary differential equation (ODE) models take the general form
\begin{equation}
  \frac{dx}{dt} = f(x(t), p),
  \label{eq:odefn}
\end{equation}
in which $x(t)$ represents the concentrations of the state variables, such as mRNA and protein concentrations, $f$ contains information on the production, degradation, and reactivity of the states, and $p$ are the kinetic parameters which govern reaction kinetics.
Limit cycle models are ODE models in which the solution approaches a steady state oscillatory trajectory, satisfying the equation:
\begin{equation}
  \lim_{t \to \infty} \left[ x(t + T) - x(t) \right] = 0.
  \label{eq:limit}
\end{equation}
The period is the smallest $T > 0$ for which Eq.~\ref{eq:limit} holds.
The points on the stable limit cycle are denoted by $x^\gamma(\theta)$ with each point assigned to a value of a phase variable $\theta \in [0, 2\pi)$.
For convenience, time in Eq.~\ref{eq:odefn} can be rescaled such that the period is $2\pi$:
\begin{equation}
  \tilde{t} = \frac{2\pi}{T}t; \quad \tilde{f} = \frac{T}{2\pi}\;f; \quad \frac{dx}{d\tilde{t}} = \tilde{f}(x(\tilde{t}), p).
  \label{eq:that}
\end{equation}
The phase variable $\theta$ is therefore defined on the limit cycle as $\theta = \tilde{t}\mod 2\pi$, with $\theta = 0$ assigned to a unique and identifiable point.

@@

In this study we restrict our analysis to temporary perturbations capable of entraining an oscillatory system, excluding permanent parameter changes arising, for instance, from genetic knockout experiments.
The most simple entraining perturbation involves adding or removing components to a limit cycle system, resulting in a perturbed trajectory $x(t)$.
Here the initial conditions are determined by the strength of the perturbation and the phase at which it is applied:
\begin{equation}
  x(0) \coloneqq x^\gamma(\theta_0) + \Delta x(0).
  \label{eq:stateperturbation}
\end{equation}
This trajectory evolves according to Eq.~\ref{eq:odefn}, eventually returning to $x^\gamma$.
It is useful to express this trajectory by the deviation from the limit cycle:
\begin{equation}
  \Delta x(\tilde{t}) = x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0).
  \label{eq:delxt}
\end{equation}

In addition to perturbations directly to the state of the system, oscillators can also be perturbed by temporary changes to the parameters.
For a parameter pulse $\Delta p$ of duration $\tilde{d}$ (in radians) that ends at $\theta_0$, the oscillator deviates from the limit cycle trajectory according to:
\begin{align}
  x_{\tilde{d}}(-\tilde{d}) &= x^\gamma(\theta_0 - \tilde{d}) \\
  \frac{dx_{\tilde{d}}}{d\tilde{t}} &= \tilde{f}(x_{\tilde{d}}(\tilde{t}), p + \Delta p) \label{eq:ode_pert}.
\end{align}
The pulse trajectory $x_{\tilde{d}}(\tilde{t})$ is then integrated from $\tilde{t} = -\tilde{d} \to 0$, at which point the pulse is removed.
A temporary parameter pulse is therefore equivalent to a state perturbation, but with the perturbation at $t = 0$ defined by:
\begin{equation}
  \Delta x(0) = x_{\tilde{d}}(0) - x^\gamma(\theta_0).
\end{equation}
A schematic depicting an example perturbed and reference trajectory is shown in Fig. 1A.

The response of the system to very small perturbations is more efficiently calculated using ODE sensitivity analysis \cite{Rabitz1983}.
The sensitivity matrix,
\begin{equation}
  S(t) = \frac{dx(t)}{dx(0)} = \lim_{\Delta x(0) \to 0}\frac{\Delta x(t)}{\Delta x(0)},
  \label{eq:senslimit}
\end{equation}
can be calculated directly by integrating
\begin{equation}
  \frac{d}{dt} S(t)  = \frac{df(x(t),p)}{dx}\; S(t)
  \label{eq:odesens}
\end{equation}
with $S(0) = I$.

\subsection*{Phase response curves}
Quantifying phase changes following a perturbation has been well studied and is particularly relevant for models of circadian rhythms \cite{Kramer1984, Taylor2008a}.
A phase response curve (PRC) maps the change in phase resulting from the same perturbation applied at each initial phase.
Infinitesimal PRCs, the derivative of the phase change with respect to the perturbation, can be defined for state and parameter-impulse perturbations \cite{Taylor2008a}.
Methods for efficiently calculating these quantities using ODE sensitivity analysis have been developed \cite{Taylor2008a}, with the important result that the parameter- and state-impulse PRCs can be related by the Jacobian matrix:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp}.
  \label{eq:pPRCequiv}
\end{equation}
This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.

\subsection*{Phase-diffusion model}
Large populations of oscillators are typically described using phase-only models \cite{Rougemont2006}, in which the state of each oscillator is represented only by its phase, $\theta$.
The synchrony of the population can be modeled using a probability density function $p(\theta, \tilde{t})$ that describes the probability of finding an oscillator at each phase \cite{Kuramoto1984}.
The usefulness of probability density functions in describing the phase and amplitude responses of populations of circadian cells has been previously shown \cite{Ukai2007}.
As with all probability density functions,
\begin{equation}
  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1.
\end{equation}
The shape of $p(\theta, \tilde{t})$ changes as the cells advance in time.
Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population \cite{Teramae2004}.
For a infinite population of oscillators, these effects are well-described by a Fokker-Plank equation \cite{Stein1965}:
\begin{equation}
  \frac{\partial p}{\partial \tilde{t}} = \frac{\partial p}{\partial \theta} + d\frac{\partial^2 p}{\partial \theta^2}.
  \label{eq:pde}
\end{equation}
Due to the rescaling of $t$, the mean period of the population is $2\pi$.
Here, the $\nicefrac{\partial p}{\partial \theta}$ term, analogous to convection, describes the mean oscillatory period, while the $\nicefrac{\partial^2 p}{\partial \theta^2}$ term describes the diffusion of phases across $[0, 2\pi)$.
The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data \cite{Rougemont2007}.
Eq.~\ref{eq:pde} has periodic boundary conditions, with initial condition $\phi(\theta)$ as the phase population at $t=0$:
\begin{align}
  \text{BCs:}\quad p(0, \tilde{t}) &= p(2\pi, \tilde{t}) \\
  \frac{\partial p}{\partial \theta}(0, \tilde{t}) &= \frac{\partial p}{\partial \theta}(2\pi, \tilde{t}) \\
  \text{IC:}\quad p(\theta, 0) &= \phi(\theta).
  \label{eq:pde_ic}
\end{align}
The solution of Eqs.~\ref{eq:pde}-\ref{eq:pde_ic} is well-characterized, with $p(\theta, \tilde{t})$ evolving in time as the convolution of the initial conditions with a wrapped normal distribution with mean $\tilde{t}$ and standard deviation $\sqrt{2d\tilde{t}}$ \cite{Chirikjian2009}:
\begin{equation}
  p(\theta, \tilde{t}) = \phi(\theta) * \mathcal{WN}(\theta; \tilde{t},
  \sqrt{2\, d\, \tilde{t}}),
\end{equation}
in which the wrapped normal distribution \cite{Mardia2009} is defined as
\begin{equation}
  {\cal WN}(\theta; \mu, \sigma) =
  \frac{1}{\sigma\sqrt{2\pi}}\sum_{k=-\infty}^\infty \exp\left[\frac{-(\theta
  - \mu + 2\pi k)^2}{2\sigma^2}\right].
  \label{eq:wrapped_normal}
\end{equation}
Since the convolution of two normal distributions is also a normal distribution, it is efficient when possible to describe $\phi(\theta)$ as a normal distribution with mean $\mu_0$ and standard deviation $\sigma_0$, such that $p(\theta, \tilde{t})$ can be found analytically through
\begin{equation}
  p(\theta, \tilde{t}) = \mathcal{WN}(\theta; \mu_0 + \tilde{t},
  \sqrt{\sigma_0^2 + 2d^2\tilde{t}^2}).
\end{equation}

@@

Following a perturbation to a single cell, the path that the perturbed trajectory $x(t)$ takes in returning to the limit cycle will have a different amplitude than the unperturbed trajectory.
While such an amplitude change will only have a finite duration, it plays an important role when perturbations are repeatedly received by the clock, such as a peripheral oscillator entrained to daily metabolic stimuli.
To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times.
Since $x(t)$ approaches the reference as $t \to \infty$, the means of both trajectories are equal and can be calculated by
\begin{equation}
  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta.
  \label{eq:mu}
\end{equation}
The amplitude change metric is defined as
\begin{equation}
  \begin{aligned}
    \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
    &= \int_0^\infty h(t) \; dt.
  \end{aligned}
  \label{eq:ampchangedefinition}
\end{equation}
This amplitude metric was chosen over alternatives, such as peak-trough distance, for its analytical tractability.
The integrand in Eq.~\ref{eq:ampchangedefinition}, abbreviated by $h(t)$, compares the variances of the reference and perturbed trajectories.
When $h(t) > 0$, the trajectory is further from the mean than the reference, and similarly when $h(t) < 0$ the trajectory is closer to the mean than the reference.
Thus the overall amplitude change can be calculated by integrating $h(t)$ until the two trajectories converge, returning an amplitude value for each state variable.
Amplitude change for a two-dimensional oscillator is easy to visualize graphically.
In Fig.~1 B-D, the same state perturbation $\Delta x(0)$ applied at two different phases results in opposite amplitude changes, depending on whether the perturbation shifts the trajectory to the interior of the limit cycle (reduced amplitudes) or to the outside the limit cycle (increased amplitudes).
Trajectories for the second state variable, $Y(\tilde{t})$, and the corresponding integrand for the amplitude change equation, $h(\tilde{t})$, demonstrate how this transient change is quantified.

@@

We therefore derive an expression for the infinitesimal ARC, defined as

@@

  \frac{dA}{dx} \coloneqq \lim_{\Delta x(0) \to 0} \frac{\Delta A\left(x(\tilde{t}),\ x^\gamma(\tilde{t} + \theta_0 + \Delta\theta)\right)}{\Delta x(0)} = \int_0^\infty \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} \; dt.

@@

  O(\Delta\theta^2).

@@

Simplifying the integrand in Eq.~\ref{eq:sARC},

@@

    2(x^\gamma(t_\theta) - \mu)\right) \right].

@@

  &= 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right).

@@

  \frac{dA}{dx} = \int_0^\infty 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right) \; dt.

@@

Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as

@@

  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p},

@@

  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}.

@@

We next briefly describe how to calculate the PRC and ARC for a phase-density model.
These approaches are commonly used in understanding phase models \cite{Kuramoto1984, Ukai2007}, and are presented here to match previous definitions for single cells.
A phase transition curve, $g(\theta) = \theta + \Delta\theta$, maps the phase of an oscillator before perturbation to its phase after the perturbation.
Since individual oscillators are neither created nor destroyed during the perturbation, it is possible - yet numerically difficult - to directly calculate the phase probability distribution following perturbation using the standard change of variables relation:
\begin{equation}
  \hat{p}(\theta, \tilde{t})\; dg(\theta) = p(\theta, \tilde{t})\; d\theta.
  \label{eq:pdfinversion}
\end{equation}
However, it is easier to estimate the mean and standard deviation of the perturbed population $\hat{p}(\theta, \tilde{t})$ using directional statistics \cite{Mardia2009}.
A population defined on the unit circle can be described by a complex variable $z = \rho e^{i\bar{\theta}}$, where $\bar{\theta}$ is the mean phase and $\rho$, the synchronization index, is related to the standard deviation of the population.
For $\rho = 1$, the population is clustered about one mean phase, while for $\rho = 0$ the population is evenly balanced across the unit circle.
Complex variables for the population before and after perturbation can be calculated via
\begin{align}
  z &\coloneqq \int_0^{2\pi} e^{i\theta} p(\theta, \tilde{t}) \; d\theta \label{eq:zbar}\\
  \hat{z} &\coloneqq  \int_0^{2\pi} e^{i\theta} \hat{p}(\theta, \tilde{t}) \; d\theta =
  \int_0^{2\pi} e^{i g(\theta)} p(\theta, \tilde{t}) \; d\theta.
  \label{eq:zhat}
\end{align}
In Eq.~\ref{eq:zhat}, we avoid calculating the perturbed population explicitly by instead integrating over the new phases at the prior population density function.
Population-level amplitude and phase responses can therefore be calculated by
\begin{align}
  \Delta \bar{\theta} &= \angle z - \angle \hat{z} \\
  \Delta \rho &= |z| - |\hat{z}|.
  \label{eq:popampchange}
\end{align}
Population-level PRCs and ARCs can be tabulated by solving Eqs.~\ref{eq:zbar}-\ref{eq:popampchange} for populations $p(\theta, \tilde{t})$ with different mean phases.
It is important to note that the ARC at the population level strongly depends on the slope of the PRC, as has been shown previously \cite{Ukai2007}.

@@

Assuming each oscillator in the population follows the dynamics described by $x^\gamma(\theta)$, the unperturbed mean population-level expression, $\bar{x}(\tilde{t})$, can be found by taking the weighted average of the expression level over the current population:
\begin{equation}
  \bar{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) p(\theta, \tilde{t}) \; d\theta.
  \label{eq:xbar}
\end{equation}

Phase-diffusion models can explain why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, a result that has been shown experimentally \cite{Welsh2004} and computationally \cite{Rougemont2007}.
Our method similarly demonstrates exponential decay: for the idealized system $x^\gamma(\theta) = \cos(\theta)$ starting from a synchronized state:
\begin{align}
  \bar{x}(\tilde{t}) &= \frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty} \cos(x) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx \\
  &= \Re\left[\frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty} \exp(ix) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx\right] \\
    &= \Re\left[e^{(i - d)\tilde{t}}\right]\\
    &= e^{-d\tilde{t}} \cos(\tilde{t}).
    \label{eq:expsin}.
\end{align}
Due to the smoothing effect of phase diffusion, higher frequency sinusoidal components of the limit cycle are damped faster than lower frequency components, resulting in an exponentially decaying sinusoids even for limit cycles that are not sinusoidal in sufficiently disperse populations.
The analytical result in Eq.~\ref{eq:expsin} allows us to easily estimate the phase diffusivity parameter in Eq.~\ref{eq:pde} by simply fitting an exponentially damped sinusoid to detrended bioluminescence data.
